---
title: "From Threat Intelligence to Defense Cleverness: A Data Science Approach"
author: "Alex Pinto"
date: "February 2nd, 2015"
output: html_document
---

This is the companion R Markdown document to the following presentations that
were delivered in Winter 2015:

- SANS CTI Summit 2015 

Adding the TIQ-TEST functions
```{r, message=FALSE}
## Some limitations from not being an R package: Setting the Working directory
tiqtest.dir = file.path("..", "tiq-test")
current.dir = setwd(tiqtest.dir)
source("tiq-test.R")

## Setting the root data path to where it should be in this repo
.tiq.data.setRootPath(file.path(current.dir, "data"))
```

## Acessing the data using TIQ-TEST

We have roughly 2 months of data available on this public dataset:
```{r, message=FALSE}
print(tiq.data.getAvailableDates("raw", "public_outbound"))
print(tiq.data.getAvailableDates("raw", "public_inbound"))
```

This time, we also have a couple of private data feeds over some of this time,
but the information in them cannot be shared publicly as a part of this release:

```{r, message=FALSE}
if (tiq.data.isDatasetAvailable("raw", "private1")) {
  print(tiq.data.getAvailableDates("raw", "private1"))
}
```

This is an example of "RAW" (not enriched) outbound data imported from combine output
```{r, message=FALSE}
outbound.ti = tiq.data.loadTI("raw", "public_outbound", "20141101")
outbound.ti[, list(entity, type, direction, source, date)]
```

We can use the same `loadTI` function to also gather the enriched datasets:
```{r, message=FALSE}
enrich.ti = tiq.data.loadTI("enriched", "public_outbound", "20141101")
enrich.ti = enrich.ti[, notes := NULL]
tail(enrich.ti)
```

This specific outbound dataset has the following sources included:

```{r, message=FALSE}
outbound.ti = tiq.data.loadTI("raw", "public_outbound", "20141101")
unique(outbound.ti$source)
```

We can do the same for the inbound data we have to see the sources we have available:
```{r, message=FALSE}
inbound.ti = tiq.data.loadTI("raw", "public_inbound", "20141101")
unique(inbound.ti$source)
```

Here are some results of running the Novelty test on the inbound data:

```{r, fig.height=10, fig.width=12, fig.align='center', warning=FALSE}
inbound.novelty = tiq.test.noveltyTest("public_inbound", "20141001", "20141130", 
                                			 select.sources=c("alienvault", "blocklistde", 
                                                 				"dshield", "charleshaley"),
																			 .progress=FALSE)
tiq.test.plotNoveltyTest(inbound.novelty, title="Novelty Test - Inbound Indicators")
```

And results running on the outbound data:

```{r, fig.height=10, fig.width=12, fig.align='center', warning=FALSE}
outbound.novelty = tiq.test.noveltyTest("public_outbound", "20141001", "20141130", 
                                        select.sources=c("alienvault", "malwaregroup", 
                                                         "malcode", "zeus"),
																			 .progress=FALSE)
tiq.test.plotNoveltyTest(outbound.novelty, title="Novelty Test - Outbound Indicators")
```

## Overlap Test

This is an example of applying the Overlap Test to our inbound dataset
```{r, fig.height=10, fig.width=10, fig.align='center'}
  overlap = tiq.test.overlapTest("public_inbound", "20141101", "enriched", 
                                 select.sources=NULL)
  overlap.plot = tiq.test.plotOverlapTest(overlap, title="Overlap Test - Inbound Data - 20141101")
  print(overlap.plot)
```

Similarly, an example applying the Overlap Test to the outbound dataset
```{r, fig.height=10, fig.width=10, fig.align='center'}
  overlap = tiq.test.overlapTest("public_outbound", "20141101", "enriched", 
                                 select.sources=NULL)
  overlap.plot = tiq.test.plotOverlapTest(overlap, title="Overlap Test - Outbound Data - 20141101")
  print(overlap.plot)
```

With the population data we can generate some plot to compare the top quantities
of reported IP addresses on a specific date by Country

```{r, fig.height=10, fig.width=10, fig.align='center'}
  outbound.pop = tiq.test.extractPopulationFromTI("public_outbound", "country", 
                                                  date = "20141111",
                                                  select.sources=NULL, split.ti=F)
  inbound.pop = tiq.test.extractPopulationFromTI("public_inbound", "country", 
                                                 date = "20141111",
                                                 select.sources=NULL, split.ti=F)

  complete.pop = tiq.data.loadPopulation("mmgeo", "country")
  tiq.test.plotPopulationBars(c(inbound.pop, outbound.pop, complete.pop), "country")
```

```{r, fig.height=10, fig.width=12, fig.align='center'}
outbound.aging = tiq.test.agingTest("public_outbound", "20141001", "20141130")
tiq.test.plotAgingTest(outbound.aging, title="Aging Test - Outbound Data")
```

```{r, fig.height=10, fig.width=12, fig.align='center'}
inbound.aging = tiq.test.agingTest("public_inbound", "20141001", "20141130")
tiq.test.plotAgingTest(inbound.aging, title="Aging Test - Inbound Data")
```

```{r, fig.height=10, fig.width=12, fig.align='center'}
outbound.aging = tiq.test.agingTest("public_outbound", "20141001", "20141130", type="enriched",
																		split.ti=F)
tiq.test.plotAgingTest(outbound.aging, title="Aging Test - Outbound Data")
```

```{r, fig.height=10, fig.width=12, fig.align='center'}
private.aging = tiq.test.agingTest("private1", "20141001", "20141130", type="enriched",
                                    split.ti=F)
tiq.test.plotAgingTest(private.aging, title="Aging Test - Private Outbound Data", density.limit=0.7)
```

```{r, fig.height=10, fig.width=10, fig.align='center'}
  outbound.pop = tiq.test.extractPopulationFromTI("public_outbound", "country", 
                                                  date = "20141110",
                                                  select.sources=NULL, split.ti=F)
  private.pop = tiq.test.extractPopulationFromTI("private1", "country", 
                                                 date = "20141110",
                                                 select.sources=NULL, split.ti=F)

  tiq.test.plotPopulationBars(c(private.pop, outbound.pop), "country", title="Comparing Private1 and Public Feeds on 20141110")
```


```{r, fig.height=10, fig.width=12, fig.align='center'}
private.novelty = tiq.test.noveltyTest("private1", "20141001", "20141130", split.tii=F,
																			 .progress=FALSE)
tiq.test.plotNoveltyTest(private.novelty)
```

```{r, fig.height=10, fig.width=12, fig.align='center'}
outbound.novelty = tiq.test.noveltyTest("public_outbound", "20141001", "20141130",split.tii=F,
																			 .progress=FALSE)
tiq.test.plotNoveltyTest(outbound.novelty)
```

This is an example of applying the Overlap Test to our inbound dataset
```{r, fig.height=10, fig.width=10, fig.align='center'}
  overlap = tiq.test.overlapTest(c("public_outbound", "private1"), "20141101", "enriched", 
                                 split.ti=c(T,F), select.sources=NULL)
  tiq.test.plotOverlapTest(overlap, title="Overlap Test - public_outbound VS private1 - 20141101")
```

```{r, fig.height=10, fig.width=10, fig.align='center'}
  overlap = tiq.test.overlapTest(c("public_outbound", "private1"), "20141101", "raw", 
                                 split.ti=c(T,F), select.sources=NULL)
  tiq.test.plotOverlapTest(overlap, title="Overlap Test - public_outbound VS private1 - 20141101")
```

```{r, fig.height=10, fig.width=10, fig.align='center'}
uniqueTest = rbind(
	tiq.test.uniquenessTest("public_inbound", "20141001","20141001", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_inbound", "20141001","20141015", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_inbound", "20141001","20141030", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_inbound", "20141001","20141129", "raw", split.tii = T)
)

uniqueTest[count == 1]
tiq.test.plotUniquenessTest(uniqueTest, title="Uniqueness Test - Inbound Data")
```

```{r, fig.height=10, fig.width=10, fig.align='center'}
uniqueTest = rbind(
	tiq.test.uniquenessTest("public_outbound", "20141001","20141001", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_outbound", "20141001","20141015", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_outbound", "20141001","20141030", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_outbound", "20141001","20141129", "raw", split.tii = T)
)

uniqueTest[count == 1]
tiq.test.plotUniquenessTest(uniqueTest, title="Uniqueness Test - Outbound Data")
```

```{r, fig.height=10, fig.width=10, fig.align='center'}
uniqueTest = rbind(
	tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141001", "raw", split.tii = c(T,F)),
	tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141015", "raw", split.tii = c(T,F)),
	tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141030", "raw", split.tii = c(T,F)),
	tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141129", "raw", split.tii = c(T,F))
)

uniqueTest[count == 1]
tiq.test.plotUniquenessTest(uniqueTest, title="Uniqueness Test (raw) - Private Data vs. Outbound Data")
```


```{r, fig.height=10, fig.width=10, fig.align='center'}
uniqueTest = rbind(
	tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141001", "enriched", split.tii = c(T,F)),
	tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141015", "enriched", split.tii = c(T,F)),
	tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141030", "enriched", split.tii = c(T,F)),
	tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141129", "enriched", split.tii = c(T,F))
)

uniqueTest[count == 1]
tiq.test.plotUniquenessTest(uniqueTest, title="Uniqueness Test (enriched) - Private Data vs. Outbound Data")
```
