---
title: "From Threat Intelligence to Defense Cleverness: A Data Science Approach"
author: "Alex Pinto"
date: "February 2nd, 2015"
output: html_document
---

This is the companion R Markdown document to the following presentations that
were delivered in Winter 2015:

* nbtcon 2014: "From Threat Intelligence to Defense Cleverness: A Data Science Approach"
* SANS CTI Summit 2015: "From Threat Intelligence to Defense Cleverness: A Data Science Approach"

This markdown file calculates the outputs and charts that are used on the presentations
using the test data available. It is published in Rpubs [here](http://rpubs.com/alexcpsec/tiq-test-Winter2015)

It should provide enough examples for usage of the tools implemented at TIQ-test.
Please review our [github repository page](https://github.com/mlsecproject/tiq-test),
report bugs and suggest features!

## Adding the TIQ-TEST functions
```{r, message=FALSE}
## Some limitations from not being an R package: Setting the Working directory
tiqtest.dir = file.path("..", "tiq-test")
current.dir = setwd(tiqtest.dir)
source("tiq-test.R")

## Setting the root data path to where it should be in this repo
.tiq.data.setRootPath(file.path(current.dir, "data"))
```

## Acessing the data using TIQ-TEST

We have roughly 2 months of data available on this public dataset:
```{r, message=FALSE}
print(tiq.data.getAvailableDates("raw", "public_outbound"))
print(tiq.data.getAvailableDates("raw", "public_inbound"))
```

This time, we also have a private data feeds over the time period,
but the information in them cannot be shared publicly as a part of this release.
If you are reproducing this at your own environemnt, you will not be able to 
recreate some of the outputs below:

```{r, message=FALSE}
if (tiq.data.isDatasetAvailable("raw", "private1")) {
  print(tiq.data.getAvailableDates("raw", "private1"))
} else {
	print("Sorry, private1 dataset is not available.")
}
```

# Data manipulation demonstration using TIQ-test

This is an example of "RAW" (not enriched) outbound data imported from combine output
```{r, message=FALSE}
outbound.ti = tiq.data.loadTI("raw", "public_outbound", "20141101")
outbound.ti[, list(entity, type, direction, source, date)]
```

We can use the same `loadTI` function to also gather the enriched datasets:
```{r, message=FALSE}
enrich.ti = tiq.data.loadTI("enriched", "public_outbound", "20141101")
enrich.ti = enrich.ti[, notes := NULL]
tail(enrich.ti)
```

This specific outbound dataset has the following sources included:

```{r, message=FALSE}
outbound.ti = tiq.data.loadTI("raw", "public_outbound", "20141101")
unique(outbound.ti$source)
```

We can do the same for the inbound data we have to see the sources we have available:
```{r, message=FALSE}
inbound.ti = tiq.data.loadTI("raw", "public_inbound", "20141101")
unique(inbound.ti$source)
```

# Novelty Test examples

Here are some results of running the Novelty test on the inbound data:

```{r, fig.height=10, fig.width=12, fig.align='center', warning=FALSE}
inbound.novelty = tiq.test.noveltyTest("public_inbound", "20141001", "20141130", 
                                			 select.sources=c("alienvault", "blocklistde", 
                                                 				"dshield", "charleshaley"),
																			 .progress=FALSE)
tiq.test.plotNoveltyTest(inbound.novelty, title="Novelty Test - Inbound Indicators")
```

And results running on the outbound data:

```{r, fig.height=10, fig.width=12, fig.align='center', warning=FALSE}
outbound.novelty = tiq.test.noveltyTest("public_outbound", "20141001", "20141130", 
                                        select.sources=c("alienvault", "malwaregroup", 
                                                         "malcode", "zeus"),
																			 .progress=FALSE)
tiq.test.plotNoveltyTest(outbound.novelty, title="Novelty Test - Outbound Indicators")
```

We can analyze the `public_outbound` dataset as a single unit as well, in order to
compare it with other repositories:

```{r, fig.height=10, fig.width=12, fig.align='center'}
outbound.novelty = tiq.test.noveltyTest("public_outbound", "20141001", "20141130",
																				split.tii=F, .progress=FALSE)
tiq.test.plotNoveltyTest(outbound.novelty)
```

The same can be done ith the inbound indicators:

```{r, fig.height=10, fig.width=12, fig.align='center'}
inbound.novelty = tiq.test.noveltyTest("public_inbound", "20141001", "20141130",
																				split.tii=F, .progress=FALSE)
tiq.test.plotNoveltyTest(inbound.novelty)
```

And with private sources we may have available:

```{r, fig.height=10, fig.width=12, fig.align='center'}
if (tiq.data.isDatasetAvailable("raw", "private1")) {
	private.novelty = tiq.test.noveltyTest("private1", "20141001", "20141130", 
																				 split.tii=F, .progress=FALSE)
	tiq.test.plotNoveltyTest(private.novelty)
} else {
	print("Sorry, private1 dataset is not available.")
}
```

## Overlap Test examples

This is an example of applying the Overlap Test to our inbound dataset
```{r, fig.height=10, fig.width=10, fig.align='center'}
overlap = tiq.test.overlapTest("public_inbound", "20141101", "enriched", 
                               select.sources=NULL)
tiq.test.plotOverlapTest(overlap, title="Overlap Test - Inbound Data - 20141101")
```

Similarly, an example applying the Overlap Test to the outbound dataset
```{r, fig.height=10, fig.width=10, fig.align='center'}
overlap = tiq.test.overlapTest("public_outbound", "20141101", "enriched", 
                               select.sources=NULL)
tiq.test.plotOverlapTest(overlap, title="Overlap Test - Outbound Data - 20141101")
```

We can use this function to compare our private dataset to each different source in
our public outbound indicator libraries. This gives some interesting insight onto
data it may be using from public sources

```{r, fig.height=10, fig.width=10, fig.align='center'}
overlap = tiq.test.overlapTest(c("public_outbound", "private1"), "20141101", "enriched", 
                               split.ti=c(T,F), select.sources=NULL)
tiq.test.plotOverlapTest(overlap, title="Overlap Test - public_outbound VS private1 - 20141101")
```

## Population Test Chart examples

With the population data we can generate some plot to compare the top quantities
of reported IP addresses on a specific date by Country

```{r, fig.height=10, fig.width=10, fig.align='center'}
outbound.pop = tiq.test.extractPopulationFromTI("public_outbound", "country", 
                                                date = "20141111",
                                                select.sources=NULL, split.ti=F)
inbound.pop = tiq.test.extractPopulationFromTI("public_inbound", "country", 
                                               date = "20141111",
                                               select.sources=NULL, split.ti=F)

complete.pop = tiq.data.loadPopulation("mmgeo", "country")
tiq.test.plotPopulationBars(c(inbound.pop, outbound.pop, complete.pop), "country")
```

We can use the same to compare our agregated outbound indicators against the
private dataset we have:

```{r, fig.height=10, fig.width=10, fig.align='center'}
if (tiq.data.isDatasetAvailable("enriched", "private1")) {
	outbound.pop = tiq.test.extractPopulationFromTI("public_outbound", "country", 
	                                                date = "20141110",
	                                                select.sources=NULL, split.ti=F)
	private.pop = tiq.test.extractPopulationFromTI("private1", "country", 
	                                               date = "20141110",
	                                               select.sources=NULL, split.ti=F)
	
	tiq.test.plotPopulationBars(c(private.pop, outbound.pop), "country", 
															title="Comparing Private1 and Public Feeds on 20141110")
} else {
	print("Sorry, private1 dataset is not available.")
}
```

## Population Test Inference - Country data

We can use some inference tools to get a better understanding if the volume of
maliciousness we are seeing makes sense in relation to the population we consider
to be our reference population.

```{r}
outbound.pop = tiq.test.extractPopulationFromTI("public_outbound", "country", 
                                                date = "20141111",
                                                select.sources=NULL,
                                                split.ti=FALSE)
complete.pop = tiq.data.loadPopulation("mmgeo", "country")
tests = tiq.test.populationInference(complete.pop$mmgeo, 
                                     outbound.pop$public_outbound, "country",
                                     exact = TRUE, top=10)

# Whose proportion is bigger than it should be?
tests[p.value < 0.05/10 & conf.int.end > 0][order(conf.int.end, decreasing=T)]

# Whose is smaller?
tests[p.value < 0.05/10 & conf.int.start < 0][order(conf.int.start, decreasing=F)]

# And whose is the same? ¯\_(ツ)_/¯
tests[p.value > 0.05/10]
```

This tool also enables us to do trend comparison between the same TI groupings 
from different days or between different groupings. A suggested usage is comparing
the threat intelligence feeds you have against the population of confirmed attacks
or firewall blocks you have in your environment.

```{r}
outbound.pop2 = tiq.test.extractPopulationFromTI("public_outbound", "country", 
                                                 date = "20141112",
                                                 select.sources=NULL,
                                                 split.ti=FALSE)
tests = tiq.test.populationInference(outbound.pop$public_outbound, 
                                     outbound.pop2$public_outbound, "country",
                                     exact = F, top=10)

# Whose proportion is bigger than it should be?
tests[p.value < 0.05/10 & conf.int.end > 0][order(conf.int.end, decreasing=T)]

# Whose is smaller?
tests[p.value < 0.05/10 & conf.int.start < 0][order(conf.int.start, decreasing=F)]

# And whose is the same? ¯\_(ツ)_/¯
tests[p.value > 0.05/10]
```

## Aging Test examples

The aging test will try to identify how long a specific indicator has lived in a
threat feed. As with other tests, like the population and novelty, you are able
to measure this information on aggregate of all your subgroups or separately.

Here is it run against the whole dataset on the Outbound indicators, as they are
separated out on subgroups:

```{r, fig.height=10, fig.width=12, fig.align='center'}
outbound.aging = tiq.test.agingTest("public_outbound", "20141001", "20141130")
tiq.test.plotAgingTest(outbound.aging, title="Aging Test - Outbound Data")
```

Here is it run against the whole dataset on the Inbound indicators. It is interesting
to observe how they have different distributions because of the different ways of collecting
the data:

```{r, fig.height=10, fig.width=12, fig.align='center'}
inbound.aging = tiq.test.agingTest("public_inbound", "20141001", "20141130")
tiq.test.plotAgingTest(inbound.aging, title="Aging Test - Inbound Data")
```

You can also look at it as whole thing, as to evaluate the aging of your whole
TI repository in its enriched format:

```{r, fig.height=10, fig.width=12, fig.align='center'}
outbound.aging = tiq.test.agingTest("public_outbound", "20141001", "20141130", type="enriched",
																		split.ti=F)
tiq.test.plotAgingTest(outbound.aging, title="Aging Test - Outbound Data")
```

Which allows us to compare it against the same formatted data for the private dataset:

```{r, fig.height=10, fig.width=12, fig.align='center'}
if (tiq.data.isDatasetAvailable("enriched", "private1")) {
	private.aging = tiq.test.agingTest("private1", "20141001", "20141130", type="enriched",
	                                    split.ti=F)
	tiq.test.plotAgingTest(private.aging, title="Aging Test - Private Outbound Data", density.limit=0.7)
} else {
	print("Sorry, private1 dataset is not available.")
}
```

## Uniqueness Test examples

For the Uniqueness test examples, we are calculating the absolute uniqueness of the data
on different data periods (1, 15, 30 and 60 days) to verify how this uniqueness evolves
over time. By running the tests, we see that there is not a lot of variation in the
ratio of uniqueness on inbound data:

```{r, fig.height=10, fig.width=10, fig.align='center'}
uniqueTest = rbind(
	tiq.test.uniquenessTest("public_inbound", "20141001","20141001", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_inbound", "20141001","20141015", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_inbound", "20141001","20141030", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_inbound", "20141001","20141129", "raw", split.tii = T)
)

uniqueTest[count == 1]
tiq.test.plotUniquenessTest(uniqueTest, title="Uniqueness Test - Inbound Data")
```

Neither there is a lot of variation on outbound data:

```{r, fig.height=10, fig.width=10, fig.align='center'}
uniqueTest = rbind(
	tiq.test.uniquenessTest("public_outbound", "20141001","20141001", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_outbound", "20141001","20141015", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_outbound", "20141001","20141030", "raw", split.tii = T),
	tiq.test.uniquenessTest("public_outbound", "20141001","20141129", "raw", split.tii = T)
)

uniqueTest[count == 1]
tiq.test.plotUniquenessTest(uniqueTest, title="Uniqueness Test - Outbound Data")
```

Also, adding the private data does not change the uniqueness ratios much further.
Some work had been done previously on selecting the feeds for little overlap, and
we can see that it paid off here.

```{r, fig.height=10, fig.width=10, fig.align='center'}
if (tiq.data.isDatasetAvailable("enriched", "private1")) {
	uniqueTest = rbind(
		tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141001",
														"enriched", split.tii = c(T,F)),
		tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141015",
														"enriched", split.tii = c(T,F)),
		tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141030",
														"enriched", split.tii = c(T,F)),
		tiq.test.uniquenessTest(c("public_outbound", "private1"), "20141001","20141129",
														"enriched", split.tii = c(T,F))
	)
	uniqueTest[count == 1]
	tiq.test.plotUniquenessTest(uniqueTest, title="Uniqueness Test (enriched) - Private Data vs. Outbound Data")
} else {
	print("Sorry, private1 dataset is not available.")
}
```

This finishes the analysis of this dataset. Feel free to suggest new tests and sources.
